Cognate beginnings to bilingual lexical acquisition

### 2023-07-27 index.qmd
Update website

---
title: "Trajectories"
subtitle: "Lab notes"
date: "Updated: `r format(Sys.Date(), '%Y/%m/%d')`"
authors:
  - name: Gonzalo Garcia-Castro
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    corresponding: true
    affiliations:
      - name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
        address: Ramon Trias fargas 25-27
        city: Barcelona
        state: Spain
        postal-code: 08005
  - name: Daniela S. Ávila-Varela
    orcid: 0000-0002-3518-8117
    affiliations:
      - name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
  - name: Ignacio Castillejo
    orcid: 0000-0001-7445-0416
    affiliations:
      - name: Universidad Atónoma de Madrid
        department: Departamento de Psicología
  - name: Nuria Sebastian-Galles
    orcid: 0000-0001-6938-2498
    affiliations:
      - name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
abstract: "Bilinguals face the challenging task of learning words from languages with overlapping phonologies. Floccia et al. (2018) reported larger vocabulary sizes for 24-month-old bilinguals that were learning languages that shared a greater amount of cognates (e.g., English-Dutch). The mechanisms underlying this effect remain unknown. We explore two compatible scenarios. First, we test whether cognates are learnt earlier than non-cognates. This would account for the difference in vocabulary size associated to the amount of shared cognates across languages. Second, we explore the possibility that the word-forms of one language interact with those in the other language, scaffolding the acquisition of their translation equivalents when their phonologies overlap. This mechanism, in line with the parallel activation account of bilingual speech perception, would provide a plausible explanation to why cognates are acquired ealier by bilinguals. We developed an online tool to collect parental reports of receptive and productive vocabularies from children learning Catalan and/or Spanish, and present data on receptive and productive vocabulary of bilingual toddlers aged 12 to 34 months."
format:
    html:
        toc: true
        link-external-newwindow: true
        number-sections: true
        number-offset: 0
        code-fold: true
        code-overflow: scroll
        code-line-numbers: true
        code-copy: hover
        fig-dpi: 1000
        reference-location: margin
        execute-dir: "C:/Users/U155880/Documents/trajectories/"
    pdf:
        toc: true
        number-sections: true
        colorlinks: true
        fig-dpi: 1000
bibliography: "references.bib"
csl: "manuscript/apa6.csl"
---

```{r}
#| label: setup
#| include: false
tar_load_globals() # load packages and options
tar_load_everything() # load targets as R objects

# resolve namespace conflicts
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")

# set ggplot theme and colour palette
theme_set(theme_custom()) # set custom ggplot theme
clrs <- c("#003f5c", "#58508d", "#bc5090", "#ff6361", "#ffa600")

options(
    knitr.kable.NA = "-",
    knitr.duplicate.label = "allow",
    ggplot2.ordinal.fill = clrs,
    ggplot2.ordinal.colour = clrs,
    ggplot2.discrete.fill = clrs,
    ggplot2.discrete.colour = clrs,
    ggplot2.continuous.fill = ggplot2::scale_color_gradient,
    ggplot2.continuous.colour = ggplot2::scale_color_gradient
)
```

::: {.callout-tip}
Data and materials are available at [OSF](https://osf.io/hy984/), and code is available at [GitHub](https://github.com/gongcastro/trajectories).
:::

# Methods

Data and materials are available at [OSF](https://osf.io/hy984/), and code is available at [GitHub](https://github.com/gongcastro/trajectories).

## Questionnaires

```{r}
#| label: questionnaire
#| echo: true
#| message: false
#| warning: false
data("pool", package = "bvqdev")
n_categories <- distinct(pool, semantic_category) %>% nrow()
n_items <- length(unique(pool$item))
n_item_language <- count(pool, language) %>% 
    pull(n) %>%
    set_names(c("catalan", "spanish"))
```

The questionnaire was implemented on-line using the formR platform [@arslan2020formr], and was structured in three blocks: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted of two lists of words: one in Catalan and one in Spanish. The Catalan inventory contained `r n_item_language["catalan"]` items and the Spanish inventory contained `r n_item_language["spanish"]` items. Items in one language were translation equivalents of the items in the other language (e.g., whenever *gos* \[dog\] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. When there were two acceptable translation equivalents for a given word, we included both in separate items (e.g., Catalan *acabar* \[*to finish*\] and Spanish *acabar* and *terminar*), or merged them into a single items (e.g., Spanish *mono* \[*monkey*\] and Catalan *mono/mico*. We included items from a diverse sample of `r n_categories` semantic/functional categories (see Appendix 1). For the analyses included in this study, we excluded items from the adverbs, auxiliary words, connectives, interjections and games and routines categories, so that only data from content words (nouns, adjectives, and verbs) were used.

For each word in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Some families filled a long version of the vocabulary checklists (800 translation equivalents; 800 items in Catalan, 800 items in Spanish), while others filled a shorter version (\~400 translation equivalents, \~400 items in Catalan, \~400 items in Spanish). These last families were randomly assigned to one of four different subsets of the complete list of items. These lists were designed so that each contained a representative subsample of the items from the complete list. Semantic/functional categories with less than 16 items were not divided in the short version of the questionnaire to preserve a minimum of four items per list for each category: all of their items were included in the four lists. Another subset of items that were part of the trial lists of some experiments in the lab were also included in all versions. Table 2 in Appendix 1 shows the distribution of items across questionnaire versions. We excluded from the analysis multi-word items (e.g., *barrita de cereales* \[cereal bar\]) and items that included more than one word-form (e.g., *mono / mico*). Table 3 shows the classification of items in cognates and non-cognates and their lexical frequency scores across the four lists of the inventories.

## Participants

```{r}
#| label: fig-participants-time
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 10 
#| fig-cap: "Date at which responses were collected"
participants %>% 
    group_by(time_stamp) %>% 
    summarise(n = n(), .groups = "drop") %>% 
    mutate(n = cumsum(n)) %>% 
    ggplot(aes(time_stamp, n)) +
    geom_vline(xintercept = ymd("2020-03-15"), size = 0.5) +
    annotate(geom = "text", y = 350, x = ymd("2020-03-15"), label = "COVID-19 lockdown in Spain",
             angle = 90, vjust = 1.5, hjust = 1) +
    geom_line(size = 1, colour = clrs[1]) +
    labs(x = "Date", y = "Number of responses", colour = "Dominant language") +
    scale_y_continuous(breaks = seq(0, 500, 100), limits = c(0, 500)) +
    theme(
        legend.position = "top",
        legend.title = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 20, hjust = 1),
        axis.title.x = element_blank(),
    ) +

    participants %>% 
    mutate(age = cut(floor(age), breaks = seq(7, 35, 5))) %>% 
    group_by(time_stamp, age) %>% 
    summarise(n = n(), .groups = "drop") %>% 
    group_by(age) %>% 
    mutate(n = cumsum(n)) %>% 
    ggplot(aes(time_stamp, n, colour = age)) +
    geom_vline(xintercept = ymd("2020-03-15"), size = 0.5) +
    geom_line(size = 1) +
    labs(x = "Date", y = "Number of responses", colour = "Age (months(") +
    scale_y_continuous(breaks = seq(0, 500, 50)) +
    scale_colour_manual(values = clrs) +

    plot_layout(ncol = 1) &

    scale_x_date(date_breaks = "3 month", date_labels = "%B %Y") &
    theme(
        legend.position = "top",
        legend.title = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 20, hjust = 1),
        axis.title.x = element_blank(),
    )
```

### Age

Number of months elapsed between participants' birth date and questionnaire completion.

```{r}
#| label: fig-participants-age
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Distribution of participants' ages"
participants %>% 
    mutate(age = floor(age)) %>% 
    count(age) %>% 
    ggplot() +
    aes(x = age, y = n) +
    geom_col(fill = clrs[1]) +
    geom_text(aes(label = n), size = 3.5, vjust = -1) +
    labs(
        x = "Age (months)",
        y = "# participants"
    ) +
    scale_x_continuous(breaks = seq(0, 40, 2)) +
    scale_y_continuous(limits = c(0, 55), breaks = seq(0, 55, 10)) +
    theme(
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()
    )
```

### Language exposure (Degree of Exposure)

Participant degree of exposure to the words' language ($DoE$): percentage of exposure to the language the word belonged to. For example, for a participant with 90% exposure to Catalan, and 10% to Spanish, the DoE of the Catalan word *taula* would be 90%, and the DoE for the Spanish word *mesa* would be 10%.

```{r}
#| label: fig-participants-lp
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Participant degree of exposure (DoE) to Catalan and Spanish"
participants %>% 
    pivot_longer(
        c(doe_catalan, doe_spanish),
        names_to = "language", values_to = "doe", 
        names_transform = ~str_to_sentence(str_remove(., "doe_"))
    ) %>% 
    mutate(
        doe = cut(doe, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
        id = paste0(id, " (", time, ")")
    ) %>% 
    count(language, doe) %>% 
    ggplot(aes(doe, y = n, label = n, fill = language)) +
    geom_col(colour = "white", size = 1) +
    geom_text(position = position_nudge(y = 5)) +
    facet_wrap(~language, ncol = 1) + 
    labs(
        x = "Degree of exposure to Catalan",
        y = "Number of participants",
        colour = "Language",
        fill = "Language"
    ) +
    scale_fill_manual(values = clrs[c(1, 3, 5)]) +
    theme(
        legend.position = "top",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank()
    )
```

### SES/parental education

Maximum educational attainment of the participant parent(s).

```{r}
#| label: tbl-participants-edu
#| tbl-cap: Participant exposure to the language (*DoE*) and socio-economic status (*SES*)
#| echo: true
#| message: false
#| warning: false
participants %>%
    left_join(bvq_data$logs) %>% 
    mutate(
        edu_parent = fct_explicit_na(as.factor(edu_parent)),
        doe_catalan = cut(doe_catalan, breaks = seq(0, 1, 0.2), include.lowest = TRUE)
    ) %>% 
    count(doe_catalan, edu_parent) %>% 
    right_join(
        expand_grid(
            doe_catalan = levels(.$doe_catalan),
            edu_parent = unique(bvq_data$logs$edu_parent)
        )
    ) %>% 
    rename(total = n) %>% 
    pivot_wider(names_from = edu_parent, values_from = total, values_fill = 0) %>%
    mutate(total = rowSums(cbind(.[,3:8]), na.rm = TRUE)) %>% 
    rename(Total = total) %>% 
    select(-"NA") %>% 
    gt() %>% 
    summary_rows(
        columns = Complementary:Primary, 
        fns = list(Total = ~sum(., na.rm = TRUE)),
        formatter = fmt_integer
    ) %>% 
    cols_label(doe_catalan = "DoE Catalan") %>%
    sub_missing(columns = 2:8, missing_text = "--") %>% 
    tab_spanner("Educational attainment", Secondary:Primary) %>% 
    tab_style(
        style = list(cell_borders(weight = px(2), sides = "left", color = "grey50")),
        locations = cells_body(columns = 9, rows = 1:5)
    ) %>% 
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels(columns = 2:7)
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_labels(columns = c(1, 8))
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_spanners(spanners = "Educational attainment")
    )
```

## Items

```{r}
#| label: tbl-items-summary
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| tbl-cap: "Lexical frequencies. Mean, standard error, and 95% confidence interval of lexical frequencies of items included in the Catalan and Spanish lists, reported separately for identical cognates, non-identical cognates, and non-cognates."
items <- left_join(items, select(bvqdev::pool, te, item, language, class, semantic_category))

items %>%
    unnest(list) %>% 
    group_by(language, list) %>% 
    summarise_at(
        vars(freq_zipf, n_phon, lv), 
        list(
            n = ~sum(!is.na(.)),
            mean = ~mean(., na.rm = TRUE),
            sd = ~sd(., na.rm = TRUE),
            min = ~min(., na.rm = TRUE),
            max = ~max(., na.rm = TRUE)
        )
    ) %>% 
    select(-c(n_phon_n, lv_n)) %>% 
    rename(n = freq_zipf_n) %>% 
    gt(groupname_col = "language", rowname_col = "list") %>% 
    fmt_number(matches("freq_|n_phon_mean|n_phon_sd"), decimals = 1) %>% 
    fmt_percent(matches("lv_"), decimals = 1) %>% 
    tab_spanner("Frequency (Zipf)", columns = starts_with("freq_")) %>% 
    tab_spanner("Length (Phonemes)", columns = starts_with("n_phon_")) %>% 
    tab_spanner("Phon. similarity (Levenshtein)", columns = starts_with("lv_")) %>% 
    cols_merge_uncert(col_val = freq_zipf_mean, col_uncert = freq_zipf_sd) %>% 
    cols_merge_uncert(col_val = n_phon_mean, col_uncert = n_phon_sd) %>% 
    cols_merge_uncert(col_val = lv_mean, col_uncert = lv_sd) %>% 
    cols_merge_range(col_begin = freq_zipf_min, col_end = freq_zipf_max) %>% 
    cols_merge_range(col_begin = n_phon_min, col_end = n_phon_max) %>% 
    cols_merge_range(col_begin = lv_min, col_end = lv_max) %>% 
    cols_label(
        list = "",
        n = md("*N*"),
        freq_zipf_mean = md("Mean ± SD"),
        freq_zipf_min = md("Range"),
        n_phon_mean = md("Mean ± SD"),
        n_phon_min = md("Range"),
        lv_mean = md("Mean ± SD"),
        lv_min = md("Range")
    ) %>% 
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels(everything())
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_spanners(everything())
    ) %>% 
    summary_rows(
        fns = list("Grand mean" = ~mean(.)),
        columns = ends_with("_mean")
    ) %>% 
    summary_rows(
        fns = list("N" = ~sum(.)),
        columns = n,
        formatter = fmt_integer
    )
```

### Lexical frequency

Word lexical frequency ($Frequency$): lexical frequency of the word in its corresponding language, expressed as Zipf scores [@van2014subtlex; @zipf1949human]. This variable ranges from 0 to 7, and follows and approximates a normal distribution, with most values in a corpus ranging from 3 to 5 points. Lexical frequencies were extracted from the English corpora of the [CHILDES](https://childes.talkbank.org/) database using the [{childesr}](https://langcog.github.io/childes-db-website/) R package. Available words in the corpora were then mapped to their Catalan and Spanish translations. Responses to words with missing lexical frequency score were excluded from analyses of the total number of items).

```{r}
#| label: fig-items-freq
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Distribution of lexical frequency across items"
items %>% 
    ggplot() +
    aes(
        x = language,
        y = freq_zipf,
        fill = language,
        colour = language
    ) +
    stat_slab(slab_colour = "white", slab_size = 1, alpha = 0.5, scale = 0.5) +
    stat_dots(side = "left", scale = 0.5, position = position_nudge(x = -0.05)) + 
    stat_pointinterval(
        position = position_dodge(width = 0.1),
        point_size = 2.5,
        colour = "black"
    ) +
    labs(
        x = "Frequency (Zipf)",
        y = "Probability density",
        fill = "Language",
        shape = "Language",
        colour = "Language"
    ) +
    theme(
        legend.position = "none",
        legend.title = element_blank()
    )
```

### \# Phonemes

Word length ($Phonemes$), computed as the number of phonemes in the phonological transcription of the word-form in [SAMPA](https://en.wikipedia.org/wiki/SAMPA) format.

```{r}
#| label: fig-items-nphon
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Distribution of number of phonemes across items"
items %>%
    count(language, n_phon) %>% 
    ggplot() +
    aes(
        x = n_phon,
        y = n,
        fill = language,
        colour = language,
        shape = language
    ) +
    facet_wrap(~language, ncol = 1) +
    geom_col() +
    geom_text(aes(label = n), position = position_nudge(y = 5), colour = "black") +
    labs(
        x = "# Phonemes",
        y = "Number of words",
        fill = "Language",
        shape = "Language",
        colour = "Language"
    ) +
    scale_x_continuous(breaks = 1:15) +
    theme(
        legend.position = "none",
        axis.ticks.x = element_line(),
        legend.title = element_blank()
    )
```

### Phonological similarity (Levenshtein)

Phonological similarity between translation equivalents was computed as the normalised Levenshtein similarity between the word-form and its translation in the other language [@levenshtein1966binary]. This score is calculated by first calculating the Levenshtein distance between the two transcriptions (number of insertions, deletions or replacements needed for the shortest transcription to become identical to the longer transcription), then dividing the resulting value by the length of the longest transcription, and finally subtracting this value from 1. This results in a proportion that indicates how much the two phonological transcriptions of the translation equivalent are similar to each other, ranging from 0% (no similarity at all) to 100% (both transcriptions are identical) [see @floccia2018introduction; @fourtassi2020growth; @laing2022phonological for similar approaches].

```{r}
#| label: fig-items-lv
#| echo: true
#| message: false
#| warning: false
#| fig-width: 9
#| fig-cap: "Distribution of Levenshtein similarities across items"
items %>% 
    distinct(te, .keep_all = TRUE) %>% 
    mutate(lv = cut(lv, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>% 
    count(lv) %>% 
    mutate(prop = n/sum(n)) %>% 
    ggplot(aes(x = lv, n)) +
    geom_col(fill = clrs[1], colour = "white") +
    geom_text(aes(label = paste0(n, " (", percent(prop, accuracy = 0.1), ")")),
              position = position_nudge(y = 3), size = 3.5) +
    labs(x = "Levenshtein distance", y = "Probability density") +
    theme(
        panel.grid.major.x = element_blank(),
        legend.position = "top"
    )
```

### Frequency by \# Phonemes

```{r}
#| label: fig-items-frequency-phonemes
#| echo: true
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Association betwen number of lexical frequency and number of phonemes"
items %>% 
    drop_na(freq_zipf) %>% 
    count(n_phon, freq_zipf, language) %>% 
    ggplot(aes(n_phon, freq_zipf, colour = language, fill = language)) +
    facet_wrap(~language) +
    geom_point(size = 2, alpha = 0.25) +
    geom_smooth(method = "lm", formula = "y ~ x") +
    labs(
        x = "# Phonemes", 
        y = "Lexical frequency (Zipf score)\nExtracted from SUBTLEX", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_x_continuous(breaks = seq(0, 16, 1)) +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()
    ) 
```

### Frequency by Levenshtein

```{r}
#| label: fig-items-frequency-lv
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Association betwen lexical frequency and phonological similarity"
items %>% 
    left_join(select(bvqdev::pool, item, language)) %>% 
    drop_na(freq_zipf) %>% 
    count(freq_zipf, lv, language) %>% 
    ggplot(aes(freq_zipf, lv, colour = language, fill = language, size = n)) +
    facet_wrap(~language) +
    geom_point(size = 2, alpha = 0.25) +
    geom_smooth(method = "lm", formula = "y ~ x") +
    labs(
        y = "Levenshtein similarity", 
        x = "Lexical frequency (Zipf score)\nExtracted from SUBTLEX", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_y_continuous(labels = percent) +
    theme(
        legend.position = "none",
        panel.grid.major.y = element_blank()
    )
```

### \# Phonemes by Levenshtein

```{r}
#| label: fig-items-nphon-lv
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Association betwen number of phonemes and phonological similarity"
items %>% 
    left_join(select(bvqdev::pool, item, language)) %>% 
    drop_na(freq_zipf) %>% 
    count(n_phon, lv, language) %>% 
    ggplot(aes(n_phon, lv, colour = language, fill = language, size = n)) +
    facet_wrap(~language) +
    geom_point(shape = 1, stroke = 1) +
    geom_smooth(method = "lm", formula = "y ~ x", size = 1) +
    labs(
        x = "# Phonemes", 
        y = "Levenshtein similarity", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_y_continuous(labels = percent) +
    scale_x_continuous(breaks = seq(0, 16, 1)) +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()
    ) 
```

### Language exposure-weighted lexical frequency

We created a predictor ($Exposure$) to account for the exposure rate of each child to each of the words their parents responded to, weighted by the child's exposure to the language the word belongs to. the exposure of the $i$-th child to the $j$-th word measure is the product of the lexical frequency of the word (Zipf score) and the child’s degree of exposure to the corresponding language (a proportion)

$$
\textbf{Exposure}_{ij} = \textbf{Frequency}_i \times \textbf{DoE}_{j}
$$

For instance, for a child who is reportedly exposure to Catalan 80% of the time, and to Spanish 20% of the time, the expected exposure to the word *cavall* (horse, in Catalan, with a lexical frequency of `r round(items$freq_zipf[items$item=="cat_cavall"], 2)`) would be `r round(items$freq_zipf[items$item=="cat_cavall"]*0.80, 2)`, while that of its translation to Spanish *caballo* would be `r round(items$freq_zipf[items$item=="cat_cavall"]*0.20, 2)`:

$$
\begin{aligned}
\textbf{Exposure}_{\textbf{cavall, Catalan}} &= \textbf{Frequency}_{cavall} \times \textbf{DoE}_{Catalan} = 6.14 \times 0.8 = 4.91 \\
\textbf{Exposure}_{\textbf{caballo, Spanish}} &= \textbf{Frequency}_{caballo} \times \textbf{DoE}_{Spanish} = 6.14 \times 0.2 = 1.23
\end{aligned}
$$

```{r}
#| label: fig-exposure-distribution
#| fig-cap: "Relationship between lexical frequency and language exposure-weighted lexical frequency"
#| warnings: false
#| fig-height: 7
#| fig-width: 13
df %>% 
    mutate(
        language = ifelse(grepl("cat_", item), "Catalan", "Spanish"),
        exposure = cut(exposure, breaks = seq(0, 9, 1), include.lowest = TRUE)
    ) %>% 
    count(exposure, language) %>% 
    ggplot(aes(exposure, n, fill = language)) +
    facet_wrap(~language) +
    geom_col(colour = "white") +
    labs(
        x = "Exposure-weighted frequency\n(Frequency \u00d7 DoE)",
        y = "Number of observations",
        fill ="Language"
    ) +

    df %>% 
    mutate(
        language = ifelse(grepl("cat_", item), "Catalan", "Spanish"),
        exposure = cut(exposure, breaks = seq(0, 9, 1), include.lowest = TRUE),
        freq = cut(freq, breaks = seq(0, 9, 0.25), include.lowest = TRUE)
    ) %>% 
    count(exposure, freq, language) %>%    
    ggplot(aes(x = freq, y = exposure, colour = language, size = n)) +
    facet_wrap(~language) + 
    geom_point() +
    labs(
        x = "Frequency (Zipf score)",
        y = "Exposure-weighted frequency\n(Frequency \u00d7 DoE)",
        colour = "Language",
        fill = "Language",
        size = "Number of observations"
    ) +
    theme(panel.grid = element_blank()) +

    plot_layout(ncol = 1) &
    guides(fill = "none", colour = "none") &
    theme(panel.grid.major.x = element_blank(),
          panel.grid.minor.y = element_blank())
```

## Data analysis

### Model comparison

We extracted four variables of interest for each observation (describe in previous sections):

* Word lexical frequency ($Frequency$)
* Word length ($Phonemes$)
* Phonological similarity ($Levenshtein$)
* Participant age ($Age$)
* Word exposure index ($Exposure$)

We fitted a series of nested Bayesian ordinal multilevel models to sequentially test the effects of the independent variables (number of phonemes, exposure, age, and Levenshtein) over the three levels of word learning (no, production, comprehension). In all these models participants’ and items’ individual differences were controlled by estimating different parameters for each one (different intercepts and/or slopes). This is similar to the control conducted by ANCOVAs. Thus, independent variable effects account for collective differences in word learning (unexplained by participants’ or items’ individual differences).

We initially fitted a base model (`model_fit_0`) that only included the predictors $Age$ (`age_std`), $Phonemes$ (`n_phon_std`), along with random intercepts and slopes of all predictors by $Participant$ (`id`) and $Item$ (`te`), and all correlation parameters. We then expanded this model by adding the main effect of $Exposure$ (`exposure`) (`model_fit_1`), and then $Levenshtein$ (`lv_std`) (`model_fit_2`) and their corresponding random slopes by $Participant$. Then, we added the $Exposure \times Levenshtein$ (`exposure_std:lv_std`) two-way interaction (`model_fit_3`) and its random slope by $Participant$. Finally, we added the $Age \times Exposure \times Levenshtein$ (`age_std:exposure_std:lv_std`) three-way interaction with its random slope by $Participant$, along with a random slope for the $Age \times Exposure$ (`age_std:exposure_std`) two-way interaction.

We compared the predictive performance of the models using Bayesian leave-one-out cross-validation (*LOO*), a relative fit index, via the `loo` R package. LOO consists in computing the average likelihood of each observation after estimating the model's parameters leaving that same observation out of the data set. Due to computational constraints, we performed LOO based on a random subsample of 750 samples (out of `r format(dim(model_fit_4$fit)[1], big.mark = ",")`) of the posterior distribution of the fixed effects of the model.

### Model equation

::: column-body-outset
$$\begin{aligned}
g[\mathrm{Pr}(y_{ij} \le a_s)] = \tau_s - \nu_{ij}
\end{aligned}$$
:::

$s$ is the s-th response category, $a_1 < a_2 < a_3$ are ordered response categories, $\mathrm{Pr}(y_i \le a_3) = 1$ and $\tau_s$ are threshold parameters where $\tau_1 < \tau_2 < \tau_3$.

::: column-body-outset

$$\begin{aligned}
\nu_{ij} = & \beta_0 +\beta_1\mathrm{Age}_{i} + \beta_2\mathrm{Phon}_{j} + \beta_3\mathrm{Exposure}_{ij} + \beta_4 \mathrm{Levenshtein}_{j} + & [\mbox{Main effects}]\\
&\beta_5\mathrm{(Age_{i} \cdot \mathrm{Exposure}_{ij}}) + \beta_6\mathrm{(\mathrm{Age}_{i} \cdot \mathrm{Levenshtein}_{j}}) +  \beta_7\mathrm{(Exposure_{ij} \cdot \mathrm{Levenshtein}_{j}}) + & [\mbox{Two-way interactions}]\\
&\beta_8{(\mathrm{Age}_{i} \cdot \mathrm{Exposure}_{ij} \cdot \mathrm{Levenshtein}_{j}}) + & [\mbox{Three-way interaction}]\\

&\theta_{0j}^{(\mathrm{TE})} + \theta_{1j}^{(\mathrm{TE})}\mathrm{Age}_{i} + \theta_{2j}^{(\mathrm{TE})} \mathrm{Phon}_{j} + \theta_{3j}^{(\mathrm{TE})} \mathrm{Exposure}_{ij} +  \theta_{4j}^{(\mathrm{TE})} (\mathrm{Age}_{i} \cdot \mathrm{Exposure}_{ij}) + & [\mbox{Item-level effects}]\\

&\theta_{0j}^{(\mathrm{ID})} + \theta_{1j}^{(\mathrm{ID})}\mathrm{Age}_{i} +   \theta_{2j}^{(\mathrm{ID})} \mathrm{Phon}_{j} + \theta_{3j}^{(\mathrm{ID})} \mathrm{Exposure}_{ij} + \theta_{4j}^{(\mathrm{ID})} \mathrm{Levenshtein}_{j} + \theta_{5j}^{(\mathrm{ID})} (\mathrm{Age}_{i} \cdot \mathrm{Exposure}_{ij}) & [\mbox{Participant-level effects}]
\end{aligned}$$
:::

### Model prior

::: column-body-outset
$$\begin{aligned}\beta_{0} & \sim \mathrm{Normal}(-0.25, 0.1) & [\mbox{Intercept/response category threshold}] \\
\beta_{1} &\sim \mathrm{Normal}(1, 0.1) & [\mbox{Age population-level coefficient}]\\
\beta_{2-9} &\sim \mathrm{Normal}(0, 1) & [\mbox{Rest of population-level coefficients}] \\
\theta_{0-9} & \sim \mathrm{Normal}(1, 0.1) & [\mbox{Group-level coefficient variability}]\end{aligned}$$
:::

In addition, we specified a $LKJ$ prior--$\sim LKJ(2)$ for the correlations between group-level coefficients.

::: {.panel-tabset}

#### Prior-predicted mean

```{r}
#| label: fig-prior-mean
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Expected prior predictions"
nd <- expand.grid(
    n_phon_std = 0,
    age_std = (10:36-mean(df$age))/(sd(df$age)),
    exposure_std = 0,
    lv_std = 0
)

m <- add_epred_draws(nd, model_fit_4_prior, ndraws = NULL, re_formula = NA) %>% 
    filter(.category != "No") %>% 
    pivot_wider(names_from = ".category", values_from = ".epred") %>% 
    mutate(Understands = Understands + `Understands and Says`) %>% 
    pivot_longer(
        c(Understands, `Understands and Says`),
        names_to = ".category",
        values_to = ".epred"
    ) %>% 
    mutate(
        .category = factor(
            .category,
            levels = c("Understands and Says", "Understands"),
            labels = c("Production\n(Understands and Says))", "Comprehension\n(Understands)")
        )
    )

m %>% 
    ggplot(aes(x = age_std, y = .epred)) +
    facet_wrap(~.category) +
    stat_lineribbon(size = 1, colour = NA, .width = c(0.95, 0.89, 0.67, 0.5)) +
    stat_summary(fun = "mean", geom = "line", size = 0.75) +
    scale_x_continuous(
        breaks = (seq(10, 36, 4)-mean(df$age))/(sd(df$age)),
        labels = seq(10, 36, 4)
    ) +
    scale_fill_manual(
        values = rev(clrs), 
        labels = c("95%", "89%", "67%", "50%")
    ) +
    labs(
        x = "Age (months)", 
        y = "Probability of acquisition",
        colour = "Credible interval", 
        fill = "Credible interval"
    ) +
    scale_y_continuous(labels = percent, limits = c(0, 1)) +
    theme(
        legend.position = "top",
        axis.ticks.x = element_line(colour = "black"),
        panel.grid = element_blank()
    )
```

#### Prior-predictions

```{r}
#| label: fig-prior-predictions
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Expected prior predictions"
nd <- expand.grid(
    n_phon_std = 0,
    age_std = (10:36-mean(df$age))/(sd(df$age)),
    exposure_std = 0,
    lv_std = 0
)

m <- add_predicted_draws(nd, model_fit_4_prior, ndraws = NULL, re_formula = NA)

m %>% 
    count(age_std, .prediction) %>% 
    pivot_wider(names_from = .prediction, values_from = n) %>% 
    mutate(Understands = Understands + `Understands and Says`) %>% 
    pivot_longer(c(No, Understands, `Understands and Says`), names_to = ".prediction", values_to = "n") %>%
    ggplot(aes(x = age_std, y = n, fill = .prediction)) +
    geom_col(position = position_fill()) +
    labs(
        x = "Age (months)", 
        y = "Probability of acquisition",
        colour = "Category", 
        fill = "Category"
    ) +
    scale_x_continuous(
        breaks = (seq(10, 36, 4)-mean(df$age))/(sd(df$age)),
        labels = seq(10, 36, 4)
    ) +
    scale_fill_manual(values = clrs[c(1, 3, 5)]) +
    scale_y_continuous(labels = percent) +
    theme(
        legend.position = "top",
        legend.title = element_blank(),
        axis.ticks.x = element_line(colour = "black"),
        panel.grid = element_blank(),
        panel.border = element_rect(fill = NA, colour = "black", size = 0.75),
        strip.background = element_rect(fill = "grey", colour = NA),
        plot.background = element_rect(fill = "white", colour = NA)
    )
```

:::

### Model settings

>>>>

---
title: "Cognate beginnings to bilingual lexical acquisition"
subtitle: "Lab notes"
short-title: "Cognate beginnings"
author:
  - name: Gonzalo Garcia-Castro
    corresponding: true
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    affiliations:
      - id: cbc
        name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
        address: Ramon Trias Fargas 25-27
        city: Barcelona
        region: Spain
        postal-code: "08005"
  - name: Daniela S. Avila-Varela
    orcid: 0000-0002-3518-8117
    affiliations: 
      - ref: cbc
      - id: ul
        name: Universidade de Lisboa
        department: Facultade de Letras
        address: Alameda da Universidade
        city: Lisboa
        region: Portugal
        postal-code: "1600-214"
  - name: Ignacio Castillejo
    orcid: 0000-0001-7445-0416
    affiliations: 
      - id: uam 
        name: Universidad Autónoma de Madrid
        department: Departamento de Psicología
        address: Iván Pavlov, 6
        city: Madrid
        region: Spain
        postal-code: "28049"
  - name: Nuria Sebastian-Galles
    orcid: 0000-0001-6938-2498
    affiliations: 
      - ref: cbc
keywords:
  - lexical acquisition
  - vocabulary
  - bilingualism
  - item response theory
  - bayesian
link-citations: true
editor: source
warning: false
error: false
format:
  html:
    theme: _assets/style.scss
    page-layout: article
    code-fold: true
    smooth-scroll: true
toc: true
fig-dpi: 1000
sidebar: true
---

```{r setup}
#| label: setup
#| message: false
#| warning: false
#| echo: false
# load objects -----------------------------------------------------------------
targets::tar_config_set(
    store = here::here("_targets"),
    script = here::here("_targets.R")
)

yaml_metadata <- rmarkdown::yaml_front_matter(
    here::here("manuscript", "manuscript.qmd")
)

title <- yaml_metadata$title
abstract <- yaml_metadata$abstract
thanks <- yaml_metadata$thanks

targets::tar_load(items)
targets::tar_load(participants)
targets::tar_load(responses)

library(dplyr)
library(tidyr)
library(knitr)
